Replicator-mutator and noise-perturbed replicator dynamics

### 2014-09-26 paper/03_simulations.tex
simulation results write up

The previous section showed that the \rdd can be conceptualized as a
two-step procedure. We first calculate where the replicator dynamic
would take us, and then apply diffusion. Since the latter step is the
novel addition, we should look at diffusion in isolation to understand
what the \rdd does. This is what we will do in
Section~\ref{sec:iterated-diffusion}. Section~\ref{sec:simulations}
\dots does something else \dots

@@

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
where $\impairment >= 0$ is an imprecision parameter. When $\impairment=0$ agents
perfectly discriminate between states; when $\impairment \rightarrow
\infty$ agents cannot discriminate states at
all. Figure~\ref{fig:NosofskySim} gives an impression of
Nosofsky-similarity for different parameter values. Other
formalizations of perceptual similarity are possible, including ones
that allow for different discriminability in different areas of the
state space, but we stick with Nosofsky's similarity function for the
time being, because it is a mathematically simple, yet established
notion in mathematical psychology.

@@

  \caption{Examples for Nosofsky-similarity (non-normalized) and
    derived probability of state confusion (normalized) for different
    values of $\impairment$ and $\card{\States} = 50$.}
  \label{fig:NosofskySim}
\end{figure}

@@

The example in Figure~\ref{fig:confusion-SenRec} shows that iterated
diffusion leads to sender and receiver behavior that is the same at
all choice points. After $n$ steps of diffusion an initial sender
strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial receiver
strategy will be $\Rstrat C^n$. If we assume that $\impairment > 0$ and so
$C$ is a positive row-stochastic matrix (each state could in principle
be confused as any other state, albeit with possibly a very low
probability), it follows from the Perron-Frobenius theorem that the
limit $C^\infty = \lim_{n \rightarrow \infty} C^n$ exists (it is the
Perron-projection) and all of its rows are identical. That is why, in
the limit of iterated diffusion, $C^\infty S$ has identical rows
(messages are sent with the same probability in each state), and $R
C^\infty$ likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

\begin{itemize}
\item utility function: another Nosofsky-style function
\item get data for different set sizes, impairment and utility parameters
\item 
\end{itemize}

@@

\subsubsection{Results}
In order to visualize the impact of the impairment factor on the outcome of simulations, we plot in Figure~\ref{fig:metrics-boxplots} the distribution of each of the measurements per impairment value.
%
\begin{figure}
        \centering
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Speaker-entropy-20140813-194306}
                \caption{Entropy of sender strategy.}
%                \label{fig:gull}
        \end{subfigure}
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Hearer-entropy-20140813-194306}
                \caption{Entropy of receiver strategy.}
        \end{subfigure}

        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Speaker-informativity-20140813-194306}
                \caption{Informativity of sender strategy.}
        \end{subfigure}
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Hearer-informativity-20140813-194306}
                \caption{Informativity of receiver strategy.}
        \end{subfigure}

        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Speaker-Voronoiness-20140813-194306}
                \caption{Voronoiness of sender strategy.}
        \end{subfigure}
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Hearer-Voronoiness-20140813-194306}
                \caption{Voronoiness of receiver strategy.}
        \end{subfigure}

        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Expected-utility-20140813-194306}
                \caption{Expected utility.}
        \end{subfigure}
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=\textwidth]{plots/Iterations-20140813-194306}
                \caption{Number of iterations until convergence.}
        \end{subfigure}
        \caption{Boxplots of metrics per impairment value.}
        \label{fig:metrics-boxplots}
\end{figure}
%
Most trends are quite clear.
Entropy seems to increase with higher impairment values for both sender and receiver.
This is an expected outcome, since the higher the impairment value, the more likely each player is to confuse two states, thus the more uncertain he will be in unequivocally associating a state with a message, or vice-versa.
The trend for informativity is also expected: more vague strategies convey less information, be it from the sender or the receiver perspective.

>>>>

The previous section showed that the \rdd can be conceptualized as a
two-step procedure: we first calculate where the replicator dynamic
would take us, then apply diffusion. Since the latter step is the
novel addition, we should look at diffusion in isolation to understand
what the \rdd does. This is what we will do in
Section~\ref{sec:iterated-diffusion}. Section~\ref{sec:simulations}
 then explores how diffusion and fitness-based selection interact.

@@

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
where $\impairment >= 0$ is an imprecision or indiscriminability
parameter. When $\impairment=0$ agents perfectly discriminate between
states; when $\impairment \rightarrow \infty$ agents cannot
discriminate states at all. Figure~\ref{fig:NosofskySim} gives an
impression of Nosofsky-similarity for different parameter
values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different
areas of the state space, but we stick with Nosofsky's similarity
function for the time being, because it is a mathematically simple,
yet established notion in mathematical psychology.

@@

  \caption{Examples for Nosofsky-similarity (non-normalized) and
    derived probability of state confusion (normalized) for
    $\card{\States} = 50$ and different
    values of $\impairment$.}
  \label{fig:NosofskySim}
\end{figure}

@@

The example in Figure~\ref{fig:confusion-SenRec} suggests that
iterated diffusion leads to sender and receiver behavior that is the
same at all choice points. This holds in general. After $n$ steps of
diffusion an initial sender strategy $\Sstrat$ will be $C^n \Sstrat$,
and an initial receiver strategy $\Rstrat$ will be $\Rstrat C^n$. If
we assume that $\impairment > 0$, $C$ is a positive row-stochastic
matrix (each state could in principle be confused as any other state,
albeit with possibly a very low probability). It then follows from the
Perron-Frobenius theorem that the limit $C^\infty = \lim_{n
  \rightarrow \infty} C^n$ exists (it is the Perron-projection) and
all of its rows are identical. That is why, in the limit of iterated
diffusion, $C^\infty S$ has identical rows (messages are sent with the
same probability in each state), and $R C^\infty$ likewise has
identical rows (every message is interpreted equally). All of this is
in line with the intuition that diffusion of behavior under confusion
of states iteratively equalizes behavior for similar states; if all
states can be confused for one another in principle, behavior smoothes
out entirely in the limit.

@@

The main question we should address is whether diffusion and
fitness-based replication reasonably interact, and if so, whether the
inclusion of diffusion has any noteworthy effects on the evolving
meaning of signals. Clearly, if $\impairment = 0$, the \rdd reduces to
the \rd. If $\impairment \rightarrow \infty$, the diffusion component
takes over and the \rdd reduces to the trivial iterated diffusion that
we looked at in the previous section. For reasonable, in-between
levels of imprecision $\impairment$, the \rdd leads to communicatively
successful, yet vague signal meaning. Moreover, suitable levels of
imprecision can have further accelerating and, therefore, unifying
effects on meaning evolution.

To explore the \rdd, we looked at numerical simulations in sim-max
games with two signals whose states were evenly spaced elements of the
unit interval. The state space always included 0 and 1. We assumed
that priors were flat: $\Pr(\state) = \Pr(\state')$ for all $\state,
\state'$. As for utility functions, we implemented another
Nosofsky-style function:
\begin{align}
  \label{eq:NosofskyUtil}
  \util(\mystate{i}, \mystate{j} \myts ; \myts \toler) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \toler^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
The tolerance parameter $\toler$ models the amount of tolerable
pragmatic slack or communicative imprecision. For This choice of
utility function is governed partly by convenience in parallel to the
similarity function, but also because it has, as we believe, the right
general properties for a communicative payoff function. Unlike
utilities that are, say, linearly or quadratically decreasing
\citep[as used
by][]{JagerMetzger2011:Voronoi-Languag,FrankeJager2010:Vagueness-Signa},
utilities that are exponentially decreasing in negative quadratic
distance can model situations where a small amount of imprecision in
communication is tolerable, whereas similarly small differences in
intolerably off interpretations matter very little, with a smooth
transition between these regimes.

\begin{figure}
  \centering
  \subfloat[][$\ns = 90$, $\toler = 0.2$, $\impairment = 0.1$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-90_tol-2_imp1_ind3501.pdf}
  }
  \subfloat[][$\ns = 90$, $\toler = 0.05$, $\impairment = 0$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-90_tol-005_imp0_ind3002.pdf}
  }

  \subfloat[][$\ns = 10$, $\toler = 0.05$, $\impairment = 0$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-10_tol-005_imp0_ind1001.pdf}
  }
  \subfloat[][$\ns = 6$, $\toler = 0.05$, $\impairment = 0.05$]{
    \includegraphics[width = 0.5\textwidth]{plots/strat_example_NS-10_tol-005_imp005_ind1201.pdf}
  }
  \caption{Example strategies under \rdd at stopping time of our simulations.}
  \label{fig:example_strats}
\end{figure}

We ran 50 trials of the \rdd, starting with randomly samples sender
and receiver strategies, for each triplet of independent parameter
values: $\ns = \card{T} \in \set{6, 10, 50, 90}$, $\impairment \in
\set{0, 0.05, 0.1, 0.2, 0.3}$, $\toler \in \set{0.05, 0.1, 0.2,
  0.3}$. Each trial ran for a maximum of 200 update steps of the
\rdd. This may not be enough to guarantee convergence to the eventual
attracting state, but we were interested especially in whether the
dynamics could lead to reasonable outcomes in reasonable time. A trial
was considered converged if the total amount of change between
strategies before and after the current \rdd step was smaller than
0.001. Examples for the resulting strategy pairs at stopping time are
given in Figure~\ref{fig:example_strats}.

\begin{itemize}
\item we see some interesting stuff in these strategies
\item now we look at the whole bunch
\item introduce some measures
\end{itemize}

@@

\subsubsection{Results}
In order to visualize the impact of the impairment factor on the outcome of simulations, we plot in Figure~\ref{fig:metrics-boxplots} the distribution of each of the measurements per impairment value.
%
% \begin{figure}
%         \centering
%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Speaker-entropy-20140813-194306}
%                 \caption{Entropy of sender strategy.}
% %                \label{fig:gull}
%         \end{subfigure}
%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Hearer-entropy-20140813-194306}
%                 \caption{Entropy of receiver strategy.}
%         \end{subfigure}

%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Speaker-informativity-20140813-194306}
%                 \caption{Informativity of sender strategy.}
%         \end{subfigure}
%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Hearer-informativity-20140813-194306}
%                 \caption{Informativity of receiver strategy.}
%         \end{subfigure}

%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Speaker-Voronoiness-20140813-194306}
%                 \caption{Voronoiness of sender strategy.}
%         \end{subfigure}
%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Hearer-Voronoiness-20140813-194306}
%                 \caption{Voronoiness of receiver strategy.}
%         \end{subfigure}

%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Expected-utility-20140813-194306}
%                 \caption{Expected utility.}
%         \end{subfigure}
%         \begin{subfigure}{0.45\textwidth}
%                 \includegraphics[width=\textwidth]{plots/Iterations-20140813-194306}
%                 \caption{Number of iterations until convergence.}
%         \end{subfigure}
%         \caption{Boxplots of metrics per impairment value.}
%         \label{fig:metrics-boxplots}
% \end{figure}
%
Most trends are quite clear.
Entropy seems to increase with higher impairment values for both sender and receiver.
This is an expected outcome, since the higher the impairment value, the more likely each player is to confuse two states, thus the more uncertain he will be in unequivocally associating a state with a message, or vice-versa.
The trend for informativity is also expected: more vague strategies convey less information, be it from the sender or the receiver perspective.

### 2014-09-26 snippets/Nosofsky_similarity.Rmd
simulation results write up

simDF = data.frame(distance = rep(states,2*length(alphas)),
                   alpha = factor(rep(rep(alphas, each=ns),2)),
                   normalized = factor(rep(c("non-normalized", "normalized"),each=ns*length(alphas))),
                   similarity = c(as.vector(t(out)),as.vector(t(outNorm)))  )

p = ggplot(simDF, aes(x = distance, y = similarity, color=alpha)) + geom_point() +
  geom_line() + facet_grid(normalized~. , scales = "free")

>>>>

simDF = data.frame(distance = rep(states,2*length(alphas)),
                   alpha_i = factor(rep(rep(alphas, each=ns),2)),
                   normalized = factor(rep(c("non-normalized", "normalized"),each=ns*length(alphas))),
                   similarity = c(as.vector(t(out)),as.vector(t(outNorm)))  )

p = ggplot(simDF, aes(x = distance, y = similarity, color=alpha_i)) + geom_point() +
  geom_line() + facet_grid(normalized~. , scales = "free") + ylab("") + xlab("distance in physical space")
